---
title: "2023 CBC data processing"
format: html
editor: source
toc: true
---

# 1. Description

This is copied from the 2022 pipeline. 

# 2. Libraries, data, function

```{r warning = FALSE, message = FALSE}
library(tidyverse) # general manipulation
library(here) # organizing directories
library(janitor) # cleaning column names
library(readxl) # using the xlsx file
library(tanagR) # color palettes inspired by members of the Thraupidae family
```

These are the data files that are useful to have.

```{r warning = FALSE, message = FALSE}
# taxon info from NACC to organize final tallies by family
taxon_nacc <- read_csv(here::here("data", "species-lists", "NACC_list_species_2022-01-29.csv")) %>% 
  # changing common names for 2023
  mutate(common_name = case_when(
    common_name == "Yellow-olive Flycatcher" ~ "Yellow-olive Flatbill",
    common_name == "Cattle Egret" ~ "Western Cattle Egret",
    TRUE ~ common_name
  ))

family_levels <- taxon_nacc %>% 
  select(family) %>% 
  unique() %>% 
  mutate(family = fct_inorder(family)) %>% 
  pull(family)

# data from Murcielago
murci <- read_xlsx(here::here("data", "data-entry", "CRSR_CBC_2023_Murcielago_data_entry.xlsx"), sheet = "datos") %>% 
  clean_names() %>% 
  # attach taxonomic info
  left_join(., taxon_nacc, by = c("especie" = "common_name"))

# data from Sensoria-Buenos Aires route
# sensoria <- read_csv(here::here("data", "CRCA_CBC_2022_Sensoria_data_entry.csv")) %>% 
#   clean_names() %>% 
#   left_join(., taxon_nacc, by = c("especie" = "common_name"))
```

This is the function to summarize individual counts by species and sort species by family. This depends on the species column being named "especie" and the individual count column being named "numero".

```{r warning = FALSE}
count_function <- function(df) {
  
  # find columns in data frame called "especie" and "numero"
  test <- str_subset(colnames(df), "especie|numero")
  
  # if else statement
  # if columns from data frame are actually "especie" and "numero", do the summary
  if(identical(test, c("especie", "numero"))) {
    df %>% 
      group_by(especie) %>% 
      summarize(sum_ind = sum(numero)) %>% 
      ungroup() %>% 
      # attach taxonomic info
      left_join(., taxon_nacc, by = c("especie" = "common_name")) %>% 
      arrange(factor(family, levels = family_levels)) %>% 
      select(especie, sum_ind)
    # else throw an error
  } else {
    warning("Whoops! You might be missing columns 'especie' and 'numero'. Double check your data frame!")
  }
}
```

# 3. Murcielago data
```{r warning = FALSE}
murci_spp <- unique(murci$especie) %>% 
  length()

murci_tally <- count_function(murci)

head(murci_tally)

# write_csv(murci_tally, here::here("tallies", "murci-2023.csv"))
```

# 4. Sensoria data
```{r warning = FALSE}
# sensoria_spp <- unique(sensoria$especie) %>% 
#   length()
# 96 species

# sensoria_tally <- count_function(sensoria)

# head(sensoria_tally)

#write_csv(sensoria_tally, here::here("tallies", "sensoria-2022.csv"))
```

# 5. Post-processing in Excel

The spreadsheet to report route data to compilers has a column to enter count numbers. I exported tally results into .csv files, then copied those into different sheets in the reporting spreadsheet. I turned that data into a table (Control + T) then used `VLOOKUP` to automatically fill in counts in the main sheet count column. The general formula for this is `=VLOOKUP([cell with common name], [cells with table], 2, FALSE)`. One example is from the Murcielago total:
`=VLOOKUP(C185, 'murci-2023'!A$2:B$63, 2, FALSE)`.

# 6. Some visualizations

```{r}
#| fig-height: 8
#| fig-width: 10

family_summary <- function(df) {
  df %>% 
    group_by(family) %>% 
    # summarizing observations by family
    group_by(family) %>% 
    summarize(sum_obs = sum(numero)) %>% 
    ungroup() %>% 
    # set the factor levels to be in descending order by sum_obs
    arrange(sum_obs) %>% 
    mutate(family = fct_inorder(family))
}

murci_families <- family_summary(murci)

barplot_format <- function(df) {
  ggplot(df, aes(x = sum_obs, y = family)) +
    geom_col(aes(fill = family)) +
    geom_text(aes(label = sum_obs), nudge_x = max(df$sum_obs*0.02)) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, max(df$sum_obs*1.05))) +
    labs(x = "Total individuals observed", y = "Family") +
    theme_classic() +
    theme(legend.position = "none",
          axis.title = element_text(size = 20),
          title = element_text(size = 25), 
          plot.margin = margin(2, 2, 2, 2, "cm")) 
}

murci_families_barplot <- barplot_format(murci_families) +
  # use the Tangara chilensis color palette
  scale_fill_manual(values = tanagr_palette("tangara_chilensis", n = 32, discrete = FALSE)) +
  labs(title = "Murciélago observations by family")

murci_families_barplot
```

